Appendix B - Dominator tree and redundancy

### 2024-11-18 paper/appendixB.Rmd
Make appendix B to explain redundancy metrics

---
title: Appendix B - Dominator tree and redundancy
subtitle:
author: Alain Danet, Loubna El Madouri, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
fontsize: 10pt
header-includes:
   - \newcommand{\beginsupplement}{\renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
   - \usepackage{float}
   - \usepackage{multirow}
   - \usepackage{booktabs}
   - \usepackage{makecell}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand{\arraystretch}{1}
geometry: margin=2cm
bibliography: references.bib
number_sections: false
---

@@

As each trophic link does not contribute equally to the robustness of an
ecological network. In particular in a food-web, ....[Explain why the metric is
really cool]

[Implementation] An developed approach is based on the use of denomitor trees
[@bodini_using_2009; @allesina_who_2004; @allesina_functional_2009].
[...] The number of functional and redundant links are straightforward to
compute We computed the number of redundant and functional links using the
dominator tree approach [@allesina_functional_2009]. A species
$i$ is a dominator of species $j$ if $i$ mediates the connection $j$ to basal
resources such as the extinction of $i$ remove the connection of $j$ to basal
resources. It is then to identify the set of trophic links that are essential to
connect species to basal resources, i.e. functional links, and the
redundant links. This method requires a single basal root in the food-webs.

Pioneer papers explained the algorithm [@bodini_using_2009] and examples.
Replicating the algorithm could seem somewhat intimitating at the time, but the
computing the dominator tree of a food-web became straightforward. In R, it is
available in `igraph::dominator_tree()`. Below is presented the function and an
example of how to use it to compute functional and redundant links.

@@

```{r}
# Get the food-web example from Bodini et al. (2009), Appendix 1 
get_bodini_graph_appendix
out <- compute_redundancy(get_bodini_graph_appendix(), root = "R")
as.data.frame(out)
```

We used this function to reproduce the
two examples (Fig. \@ref(fig:bod1), \@ref(fig:)) showcased in the Fig.1 and Appendix
1 of @bodini_using_2009 , and the example (Fig. \@ref(fig:)) presented in the Fig.1 of
@allesina_functional_2009.

(ref:bod1-cap) bodin cap

```{r bod1, fig.cap="(ref:bod1-cap)"}
par(mfrow = c(1, 2))

bodinifig1 <- get_bodini_graph_fig1()

# Complete graph
lg <- layout_as_tree(bodinifig1, root = "root")
lg[, 2] <- -lg[, 2]
plot(bodinifig1, layout = lg, main = "Redundant & Functional\nlinks")

# Only functional links
bodinifig1_dtree <- dominator_tree(bodinifig1, root = "root")
layout <- layout_as_tree(bodinifig1_dtree$domtree, root = "root")
layout[, 2] <- -layout[, 2]
plot(bodinifig1_dtree$domtree, layout = layout, main = "Functional\nlinks only")
```

```{r bod2}
par(mfrow = c(1, 2))

g <- get_bodini_graph_appendix()

# Complete graph
lg <- layout_as_tree(g, root = "R")
lg[, 2] <- -lg[, 2]
plot(g, layout = lg, main = "Redundant + Functional\nlinks")

# Only functional links
dtree <- dominator_tree(g, root = "R")
layout <- layout_as_tree(dtree$domtree, root = "R")
layout[, 2] <- -layout[, 2]
plot(dtree$domtree, layout = layout, main = "Functional\nlinks only")
```

```{r all1}
par(mfrow = c(1, 2))

allesinafig1 <- get_allesina_graph(diag = FALSE)
lg <- layout_as_tree(allesinafig1, root = "0")
lg[, 2] <- -lg[, 2]
plot(allesinafig1, layout = lg, main = "Redundant + Functional\nlinks")

# Only functional links
allesinafig1_dtree <- dominator_tree(allesinafig1, root = "0")
layout <- layout_as_tree(allesinafig1_dtree$domtree, root = "0")
layout[, 2] <- -layout[, 2]
plot(allesinafig1_dtree$domtree, layout = layout,
  main = "Functional\nlinks only")
```

@@

redundancy_examples %>%
  select(reference, total_links,
    functional_links, redundant_links,
    prop_functional_links, prop_redundant_links
  ) %>%
  mutate(reference = paper_replacement[reference]) %>%
  rename_with(~redundancy_var_replacement()[.x]) %>%
  kable(
    digits = 2,
    row.names = FALSE,
    booktabs = T,
    caption = "Computation of the number (Nb) of functional, redundant links, and
    their proportion (Prop.) from previous published examples.",
    label = "inlarand") %>%
  collapse_rows(columns = c(1), valign = "top") %>%
  kable_styling(latex_options = "scale_down")
```

>>>>

---
title: Appendix B - Dominator tree and redundancy
subtitle:
author:
date: \today
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
fontsize: 10pt
header-includes:
   - \newcommand{\beginsupplement}{\renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
   - \usepackage{float}
   - \usepackage{multirow}
   - \usepackage{booktabs}
   - \usepackage{makecell}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand{\arraystretch}{1}
geometry: margin=2cm
bibliography: references.bib
number_sections: false
---

@@

Because species are interdependant, species loss can have cascading effects and
triger secondary extinctions [@dunne_network_2002]. So assessing the robustness
of ecological networks to species losses has been a big question in Ecology.
Previous research showed that food-web structure, such as connectance, increased
robustness [@dunne_network_2002]. The idea is that more connected food-web will
be more stable because they contain a large number of pathways convey biomass
through the food-web. However, connectance as some limitations as a robustness
metric has some limitations, because each trophic link does not contribute
equally to the robustness of an ecological network. Trophic links can be as
functional or redundant, depending on their role in transmitting biomass through
the food-web. Functional links are the set of links that is required to connect
every species to the basal resource of the food-web, all of the other being
redundant. Only functional links affect directly robustness, as their removal
will trigger secondary extinctions. The removal of redundant links in turn do
not trigger secondary extinctions, but makes the food-webs more fragile to
further extinctions [@allesina_functional_2009].

The identification of functional and redundant links is based on the use of
denomitor trees [@bodini_using_2009; @allesina_who_2004;
@allesina_functional_2009]. Dominator trees allows to identify the set of
trophic links that are essential to connect species to basal resources, i.e.
functional links, and the redundant links. Pioneer papers explained the
algorithm to compute the set of dominator trees and I refer to them for further
information [@bodini_using_2009; @allesina_who_2004; @allesina_functional_2009]
and examples. Here, we aim to show that while implementing this algorithm could
seem somewhat intimitating at the time, this algorithm is now widely
implemented. In R, it is available in `igraph::dominator_tree()`. Below we show
how functional and redundant links can be computed easily from R by replicating
three examples presented in previous studies.

@@

Here is an example with the food-web displayed in appendix of
@bodini_using_2009:

```{r, echo = TRUE}
# Get the food-web example from Bodini et al. (2009), Appendix 1 
g <- get_bodini_graph_appendix()
# compute reduncancy
out <- compute_redundancy(g, root = "R")
as.data.frame(out)
```

```{r}
vertex_color <- "white"
vertex_size  <- 40
```

We used this function to reproduce the three examples (Fig. \@ref(fig:examples)),
which were displayed in the Fig.1 and Appendix 1 of @bodini_using_2009, and
in the Fig.1 of @allesina_functional_2009. The table \@ref(tab:redun) displays
the number and the proportion of functional and redundant links.

```{r bod1}
bodinifig1 <- get_bodini_graph_fig1()

# Complete graph
p_bodinifig1_all <- ~{
  lg <- layout_as_tree(bodinifig1, root = "root")
  lg[, 2] <- -lg[, 2]
  plot(bodinifig1, layout = lg, vertex.color = vertex_color, vertex.size = vertex_size)
}

# Only functional links
bodinifig1_dtree <- dominator_tree(bodinifig1, root = "root")
p_bodinifig1_functional <- ~{
  layout <- layout_as_tree(bodinifig1_dtree$domtree, root = "root")
  layout[, 2] <- -layout[, 2]
  plot(bodinifig1_dtree$domtree, layout = layout, vertex.color = vertex_color, vertex.size = vertex_size)
}
```

```{r bod2, echo = FALSE, results = "hide"}
g <- get_bodini_graph_appendix()

# Complete graph
p_bodiniappen_all <- ~{
  lg <- layout_as_tree(g, root = "R")
  lg[, 2] <- -lg[, 2]
  plot(g, layout = lg, vertex.color = vertex_color, vertex.size = vertex_size)
}

# Only functional links
dtree <- dominator_tree(g, root = "R")
p_bodiniappen_functional <- ~{
  lg <- layout_as_tree(dtree$domtree, root = "R")
  lg[, 2] <- -lg[, 2]
  plot(
    dtree$domtree,
    layout = lg,
    vertex.color = vertex_color,
    vertex.size = vertex_size
  )
}

p_bodini <- plot_grid(
  ggdraw(p_bodinifig1_all), ggdraw(p_bodinifig1_functional),
  ggdraw(p_bodiniappen_all), ggdraw(p_bodiniappen_functional),
  labels = "auto"
)
ggsave_multiple(
  paste0("p_bodini", c(".png", ".pdf")),
  plot = p_bodini,
  path = here::here("figures"),
  scale = 2.6,
  units = c("mm"),
  width = 80,
  height = 80 * 1.5)
```

```{r all1, echo = FALSE, results = "hide"}
allesinafig1 <- get_allesina_graph(diag = FALSE)

p_allesinafig1_all <- ~{
  lg <- layout_as_tree(allesinafig1, root = "0")
  lg[, 2] <- -lg[, 2]
  plot(allesinafig1, layout = lg, vertex.color = vertex_color, vertex.size = vertex_size)
}

# Only functional links
allesinafig1_dtree <- dominator_tree(allesinafig1, root = "0")
p_allesinafig1_functional <- ~{
  layout <- layout_as_tree(allesinafig1_dtree$domtree, root = "0")
  layout[, 2] <- -layout[, 2]
  plot(
    allesinafig1_dtree$domtree,
    layout = layout,
    vertex.color = vertex_color,
    vertex.size = vertex_size
  )
}

p_examples <- plot_grid(
  ggdraw(p_bodinifig1_all), ggdraw(p_bodinifig1_functional),
  ggdraw(p_bodiniappen_all), ggdraw(p_bodiniappen_functional),
  ggdraw(p_allesinafig1_all), ggdraw(p_allesinafig1_functional),
  nrow = 3,
  labels = "auto"
)
ggsave_multiple(
  paste0("p_examples", c(".png", ".pdf")),
  plot = p_examples,
  path = here::here("figures"),
  scale = 4.0,
  units = c("mm"),
  width = 80,
  height = 80 * 1.2)
```

(ref:example-cap) Reproduced examples of categorising functional links in
food-webs using `igraph::dominator_tree()` R function. Left columns (a, c, e)
displays the complete food-web while right column displays food-web with the
functional links only (b, d, f). (a,b): reproduction of Fig. 1 of
@bodini_using_2009, (c, d): reproduction of appendix figures of @bodini_using_2009,
(e, f): reproduction of fig. 1 of @allesina_functional_2009. Note that (e, f) is
a bit different from the original, as noted by the authors, they could have
removed further redundant links (i.e. $2 \to 5$, or $3 \to 5$).

```{r examples, fig.cap = "(ref:example-cap)", out.height = "80%"}
knitr::include_graphics(here::here("figures", "p_examples.png"))
```

@@

redundancy_examples %>%
  select(reference, total_links,
    functional_links, redundant_links,
    prop_functional_links, prop_redundant_links
  ) %>%
  mutate(reference = paper_replacement[reference]) %>%
  rename_with(~redundancy_var_replacement()[.x]) %>%
  kable(
    digits = 2,
    row.names = FALSE,
    label = "redun",
    booktabs = T,
    caption = "Computation of the number (Nb) of functional, redundant links, and
    their proportion (Prop.) from previous published food-webs.",
    ) %>%
  collapse_rows(columns = c(1), valign = "top") %>%
  kable_styling(latex_options = "scale_down")
```
